---
title: "ISLR Lab 2"
author: "Suarez Rodes, Randy"
output:
  pdf_document: 
    keep_tex: no
  html_document:
    df_print: paged
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Basic Commands
Vector creatinon c() "concatenate" function. Use <- or = to declare variables or functions.
```{r  message=FALSE, warning=FALSE}
x <- c(1,3,2,5)
x
x = c(1,6,2)
x
y = c(1,4,3)
```

Adding vectors pointwise.
```{r  message=FALSE, warning=FALSE}
length(x)
length(y)
x+y
```

ls() and rm() (List of objects and remove)
```{r  message=FALSE, warning=FALSE}
ls()
rm(x,y)
ls()
```

Removing all at once
```{r  message=FALSE, warning=FALSE}
rm(list=ls())
```

The matrix function ?matrix to learn more about it (? it is used to get documentation help)
For matrix specify data, number of rows and cols.
```{r  message=FALSE, warning=FALSE}
x=matrix (data=c(1,2,3,4) , nrow=2, ncol =2)
x
```

We can specify the order of the entries using byrow, by default byrow=FALSE, hence the entries are created by column as the default setting.
```{r  message=FALSE, warning=FALSE}
x=matrix (data=c(1,2,3,4) , nrow=2, ncol =2,byrow = TRUE)
x
```

Functions work in a vectorized fashion (entry wise)

```{r  message=FALSE, warning=FALSE}
sqrt(x) #square root
x^2 #power
```


Random Data generation Ex:rnorm() Normal random variates. Random generation changes each call (use set.seed() to fix a random generator seed)
cor() correlation
```{r  message=FALSE, warning=FALSE}
x=rnorm (50)
y=x+rnorm (50, mean=50, sd=.1)
cor(x,y)

set.seed (1303)
rnorm (50)

```
mean() and var() of a vector, apply sqrt() to the output of var() to obtain standard deviation or simply use sd()
```{r message=FALSE, warning=FALSE}
set.seed(3)
y=rnorm (100)
mean(y)
var(y)
sqrt(var(y))
sd(y)
```

## Graphics

plot() function. Example plot(x,y) produces a scatterplot of the numbers in x versus the numbers
in y.

```{r message=FALSE, warning=FALSE}
x=rnorm (100)
y=rnorm (100)
plot(x,y)
plot(x,y,xlab=" this is the x-axis",ylab=" this is the y-axis", main=" Plot of X vs Y")
```

Use pdf() or jpeg() to 